When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-12-18 code/Tahkv1.Rmd
commit

\captionsetup[table]{labelformat=empty}

@@

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("stargazer")
library("rstatix")
```
\newpage

@@

This chapter examines variation in the publication of op-ed pieces by retired military officers since 1979. There are two main arguments of this chapter. The first argument is that the variation in the number, topic addressed, and tone of opinion-editorial publications published by retired military officers indicates a shift over the past 40 years in the way a number of retired military officers think about their normative roles as members of the military institution and how they interact with the public in America. The second argument is that the driver of the variation in the number, topic, and tone of op-eds authored by retired military officers is mainly the increase in political polarization that has occurred in America since the 1970s. While there are other factors that influence also responsible, such as the increase in institutional prestige of the military over the same time period, the empirical story reveals that political polarization holds more explanatory power than other factors in explaining the variation of retired military officer speech in America.

This paper proceeds in three parts. First, the paper connects the concept of military "role beliefs" to retired military officer speech. Second, the paper explores and analyzes an original data set of nearly 390 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s. This analysis ultimately demonstrates that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen both to argue about topics that fall outside of the realm of traditional national security topics, and to violate the principle of civilian control of the military when doing so. The analysis also shows that these trends have worsened as the American political climate has become increasingly polarized over the past several decades. Third, thechapter concludes by exploring the implications of these results on the military as an institution, and specifically, on its ability to remain professional in a highly polarized setting that will likely endure.

\singlespace
# How the Military Views Itself
\doublespace
## Military Role Beliefs
The concept of military "role beliefs" was introduced by J. Samuel Fitch in his book, _The Armed Forces and Democracy in Latin America._ In the book, Fitch describes military "role beliefs" as the perceptions held by military officers regarding their normative role in politics [@fitch_armed_1998, 61]. Though Fitch's focus is on militaries in Latin America, it is possible to use the concept of "role beliefs" to develop a normative starting point applicable to military officers in the United States.

Two foundations for developing a starting point for military "role beliefs" in the United States are the tandem norms of civilian control of the military, and the idea of the military acting as a non-partisan institution. Though such norms have been episodically challenged throughout American history in a variety of ways and during specific instances, these norms have nonetheless been both foundational and aspirational, particularly in the post World War Two period [@weigley_american_2001]. The norm of civilian control of the military stems directly from the US Constitution, which expressly grants control of the military (as well as specific responsibilities associated with maintaining control) to the President, Congress, and the individual states [@noauthor_us_1787-2]. Unlike the norm of civilian control, the norm of the military functioning as a non-partisan institution is not expressly spelled out in the Constitution, but rather the product of the American military's collective experiences and traditions in fighting and prevailing in the nation's wars and conflicts.

Further, a number of scholars have underscored the importance of these two norms, and the relationship between these norms and the military's ability to prevail during times of war. In his influential book, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, Samuel Huntington argues in favor of maintaining separate operating spheres between the political leaders of a democracy and the military as an institution in order to ensure civilian control of the military by rendering the military "politically sterile and neutral," or non-partisan [@huntington_soldier_1957, 84]. Huntington goes on to argue that only by such an arrangement, which he called "objective control of the military," will the military be capable of defeating significant security threats while also remaining apolitical [@huntington_soldier_1957, 83-84].

Having established the foundational norms of civilian control of the military and the military operating as an apolitical institution, the next step is to extrapolate how these norms might impact the public speech of military officers. This requires balancing these norms with the fact that military officers are nonetheless servants of the Republic who have accumulated knowledge and experiences over the course of sometimes lengthy careers (in the case of retired military officers, at least 20 and in some cases, upwards of 40 or even more years of experience), and thus, in many cases are qualified to weigh in on circumstances and situations that impact the nation. For instance, in the aftermath of the "Revolt of the Generals" in 2006, in which several retired military officers publicly criticized and called for the removal of Defense Secretary Donald Rumsfeld, Frederick Kagan argued in support of retired military officers having the freedom to weigh in policy debates, particularly those involving an active armed conflict, stating, "there is no danger to the republic in a handful of retired generals speaking their minds. There is great danger in making vital decisions about an on-going armed struggle without hearing the views of all available experts" [@kagan_let_2006]. Kagan's remarks lend support to the notion that in democracies in particular, transparent and flowing information is a precious resource that helps leaders make decisions and, more importantly, the public to gain awareness of the state's actions [need cite].

Nonetheless, there have been both formal and informal boundaries placed on various forms of political activities by military officers. Formally speaking, Department of Defense (DoD) Directive 1344.10 contains a "standing list of political behavior do's and don'ts for members of the active duty military" [@urben_wearing_2014, 1]. This list includes authorizing members of the military to engage in activities such as voting and joining a partisan or non-partisan club and to attend this club's meetings when not in uniform, and prohibits members of the military from activities such as participating in partisan fundraising, or advocating for or against a partisan cause or figure on radio or television [@department_of_defense_political_2008]. Informally, a number of recent Chairmen of the Joint Chiefs of Staff have likewise urged members of the military to refrain from engaging in overly political activities, particularly in election years, when such activity is more likely to negatively reflect on the military. For example, prior to the 2016 Presidential Election, then-Chairman of the Joint Chiefs of Staff General Joseph Dunford reminded the military that the basis upon which it is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016].

Based on these formal and informal prohibitions on the political activities conducted by military officers, it is possible to conclude that visible behaviors that are overtly partisan are to be avoided at all costs, and are inappropriate for military officers to engage in. It is also possible to extrapolate that the norm of civilian control of the military results in a corresponding prohibition on members of the military criticizing political leaders to the point that civilian control of the military is jeopardized or fails to be upheld. Where this line should be drawn in practice is not clear, however, because as Kagan has pointed out, criticisms can and often do point out blind spots in a plan or otherwise reveal shortcomings that may not otherwise have been identified. However, there is a point at which a criticism levied at a political leader by a military officer goes beyond criticism. Finally, it is possible to conclude that activities which build on the appropriate expertise and knowledge of the military officer are appropriate, provided that these activities do not violate the stipulations previously mentioned, i.e. a non-partisan military, and civilian control of the military.

## Opinion-Editorial Publication as a Political Activity?
We can now focus on the actual behavior of the military authoring op-ed pieces. Should these considered a form of political activity? I argue that it depends on what message the op-ed actually sends or is intended to send. An op-ed that strictly points out the pitfalls of the nation's combat strategy written by an experienced military officer may not be inappropriate, whereas an op-ed written by a retired military officer that argues in favor of a specific Presidential candidate is, at least by the standards put forth in this paper, entirely inappropriate. Therefore, at least two questions should be asked when examining an opinion piece authored by a retired military officer: first, does the piece violate either the norm of civilian control of the military or the military as an apolitical institution, and second, does the op-ed address an appropriate topic that a military officer is qualified to address?

Beyond these two questions, however, opinion-editorials are important because they serve as an informative barometer of how the military institution sees itself, and particularly, its role with respect to the American public. Members of the active military do not frequently write op-eds, but there have been notable exceptions over the years, perhaps none so notable as several written by then-Chairman of the Joint Chiefs of Staff General Colin Powell in the early 1990s. For instance, Powell presented the public with his view regarding the potential use of American military force in several developing "hot spots" across the world, including the Balkans, Somalia, and other places, during an October 8, 1992 piece entitled, "Why Generals Get Nervous." Published in the _New York Times_, Powell wrote in part, "decisive means and results are always to be preferred, even if they are not always possible. So you bet I get nervous when so-called experts suggest that all we need is a little surgical bombing or a limited attack. When the desired result isn't obtained, a new set of experts then comes forward with talk of a little esclalation. History has not been kind to this approach" [@powell_opinion_1992].

Powell clearly felt compelled to inform the public of how he, as the military's top officer, viewed the potential use of American military force abroad during the prevailing international context at the time. What is less known, however, is that Powell was also writing against a domestic backdrop in which there was an election coming up in less than a month that pitted the incumbent, George W. Bush, against Democratic challenger Bill Clinton. The two candidates differed with respect to foreign policy on a number of issues, and Bill Clinton was not highly respected by the military establishment, a fact that would impact the military's relationship with Clinton during his early years in office [need cite]. Whether it was proper for Powell to write such an op-ed truly depends: one could argue that the nation's top general was simply informing the public based on his years of individual experience serving in the military, in which case, from a democratic transparency standpoint, the op-ed well served the Republic; alternately, one could also argue that by writing during an election year, Powell's remarks brazenly staked the claims of the military ahead of a potential Presidential transition (which turned out to be the case), and in the process, limited the President's options for potentially using force abroad during his early days in office. The first explanation argues that from the standpoint of civil-military relations, there was no problem with Powell's opinion-editorial, while the second argues that the op-ed was nothing but problematic.

## Retired Military Officers
One might argue that the logic employed thus far is seemingly plausible for members of the active military, but that the cautionary tone and reservations regarding certain types of political activity should not apply for members of the retired military community. And indeed, there are obvious distinctions between those who are on active duty and those who have retired from the service. These distinctions are made clear in fact in portions of DoD 1344.10 [@department_of_defense_political_2008]. But the fact of the matter is that, especially for retired military officers who achieved general or flag officer rank, these retired officers sometimes carry outsized influence and impact when they engage in activities that approach on the political. During the Presidential Campaign of 2016, Retired Army General Martin Dempsey published an opinion-editorial piece in _The Washington Post_ which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the respective Republican and Democratic conventions. Dempsey wrote, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016].

Of course, speaking at a political convention on behalf of a Presidential candidate is, if not formally, at least informally the textbook definition of a political activity. Not every opinion-editorial piece supports a political candidate's agenda, or disparages the political views of another. Thus, it is important to remember that although an op-ed can be written in such a way that it is a politically inappropriate form of political behavior, it need not be. It is precisely on this point that the norms of civilian control of the military, the military as an apolitical institution, and retired members of the military expressing views that are in accord with the knowledge and skills gained during their time in uniform are appropriate.

Moreover, because multistar generals and admirals in the contemporary military possess in many cases upwards of 30 and sometimes 40 years of service, it is entirely reasonable to conclude that these high-ranking officers in particular have been privy to decisions and experiences occurring at or near the highest levels of government. Having interacted with Presidents, members of the cabinet, members of Congress, and foreign dignitaries, the experiences of retired general and flag officers are not limited to only tactical matters. Though they possess an abundance of expertise in many tactical matters that pertain to warfighting, they are also familiar with issues ranging from how wars should be fought, to future security threats looming on the horizon, to fixing the institutional culture of large organizations such as the military services, and much more.

Nonetheless, they are not experts in everything, nor should they try to be. Even General of the Army Douglas MacArthur, one of several men who ever wore the five star rank, and a hero of World War Two who later clashed with President Truman regarding American strategic goals in Asia, recognized that the military had focus areas, even if he did not personally exhibit these limitations while on active duty. Addressing the Corps of Cadets at West Point in 1962, the then-82 year old retired General told the Corps:

@@

\doublespacing
By systematically examining the op-eds written by retired military officers in major US publications over time, it is possible, I contend, to draw inferences about how the concept of a "role belief" - conceptions of one's normative role in politics - has changed among retired US military officers (if at all) [@fitch_armed_1998]. Op-eds are an interesting and appropriate medium to examine if differentiation in "role beliefs" is an object of interest. The first reason why this is the case is because as previously stated, the act of crafting and publishing an op-ed does not necessarily constitute a political activity, though depending on content, they can be. A second and releated reason is because unlike other forms of engaging with the press, such as giving interviews on television, crafting an op-ed requires the author to deliberately prepare and submit a viewpoint for publication. A deliberate and carefully-crafted message further underscores the degree to which the author believes the message he or she is sending. It is difficult to "misspeak" during or to be "misquoted" from an opinion-editorial. The author generally agrees with the message he or she is sending.

When examining op-eds authored by retired military officers over time to detect potential shifts in normative role beliefs, the key area of focus must be the degree to which retired military officers have upheld vice strayed from the norms of civilian control and non-partisanship of the military, as well as the degree to which retired officers have addressed topics that are appropriately rooted in relevant military expertise. If there are patterns of violating these norms that can be detected, it may indicate that the role beliefs of certain members of the retired military community have likewise changed. On the other hand, if there has been little deviation over time from these foundational norms, it is possible to conclude that military retirees view their normative roles in politics (role beliefs) in similar ways as previous generations of retired military officers.

#Determinants of Military Role Beliefs

Theoretically, there are at least two major determinants of military role beliefs, or the conceptions held by members of the military regarding their normative role in politics [@fitch_armed_1998]. In the most general terms, the military's involvement in politics requires both the _motive_ and the _opportunity_ to do so [@finer_man_1962]. I argue that rising political polarization serves as the _motive_ for the military's intervention in politics, while high levels of prestige, or credibility of the military institution, provide the _opportunity_ for doing so. It is important to elaborate why each of these offers the military in particular the ability to intervene in politics.^[For a richer elaboration on the mechanisms afforded by political polarization and institutional levels of prestige for intervention in politics, please see the author's theory chapter in his dissertation. The proceeding paragraphs here are intended to summarize these mechanisms.]

With respect to the first determinant, political polarization, the literature in American politics enables the development of two plausible relationships and corresponding mechanisms that connect polarization to  military role beliefs. These relationships and mechanisms further stem from a particular conception or definition of what polarization is and how it workd. One possible conception of polarization is that it marks the ideological widening between political elites, to include members of Congress or the leaders of political parties [@hare_polarization_2014]. According to this view of polarization, one possible mechanism by which the military intervenes in politics is by adopting a belief that elites from both dominant political parties have failed to advance and implement an effective governing agenda, and thus, the military steps in to do what political elites are not doing. This particular dynamic has been noted, in fact, by scholars in the comparative politics subfield describing the actions of militaries in other parts of the world [@fitch_armed_1998].

@@

Given the expectations derived from theories regarding military role beliefs, the overall purpose of opinion-editorial publications, and retired military officers as members of the military community, five distinct hypotheses are derived. These include:

\singlespace
**H1: When political polarization is high, the military (or certain military figures) will increasingly criticize civilian leadership (indicating a drop in the principle of civilian control).**

@@

\singlespace
**H2: When political polarization is low, the military (or certain military figures) will publicly weigh in on traditional national security debates (and not on domestic political concerns or issues outside of relevant military expertise).**

@@

\singlespace
**H3: During periods of high prestige and credibility, the frequency of military involvement in quasi-political activities (such as writing op-eds) will increase relative to periods of low prestige and credibility.**

@@

\singlespace
**H4: When both political polarization and high prestige are present, the military will be increasingly vocal about certain domestic ideological or partisan causes than when these factors are not present.**

@@

\singlespace
**H5: During periods of active ground combat, military elites will likely comment on the war, but with less frequency on other domestic issues.**

@@

Some caution must be taken up front so as to not overpromise and under-deliver concerning what the data is actually capable of presenting. The first caution to consider is that opinion-editorial publications by nature reflect the demand of the press to publish the viewpoints of retired military officers. Moreover, because op-eds are generally provocative and at least to some degree designed to advocate for or against a certain course of action, it is possible that op-eds may generally present somewhat contentious material. Additionally, this means that under periods of harmonious civil-military relations, there is either less of a demand from the press to publish the viewpoints of retired military officers, or less of a supply side effect for retired military officers to write about topics in the first place. Certain exogenous events, such as wars, will impact whether retired military officers speak out. A second cautionary note is that over the course of several decades, underlying societal changes - such as political polarization - have occurred that likely impact either or both the willingess of the press to publish the viewpoints of military officers, or the desire of military officers to write about certain topics in the first place. It is important that these societal-level effects be identified and, if possible, controlled for.

@@

The methodological approach used in this analysis consists of two broad steps. The first is to analyze the topics which retired military officers have been writing about, and the second is to determine, regardless of the topic addressed in a publication, whether an op-ed has violated the norms of non-partisanship or civilian control of the military. The critical question is whether there are patterns in the data that can be detected, and whether these patterns can be explained.

The data consists of 390 opinion-editorial publications and letters to the editor authored by retired military officers and published in major US newspapers from 1979-2020. I include both opinion-editorials and letters to the editor because both display a viewpoint that required the author to craft a deliberate message, and to submit the piece for consideration to be published by an editorial board at a particular newspaper. Because I am interested in messages that required the officer to craft a deliberate message, I do not include published interviews with retired military officers, nor any reprinted speeches or remarks that a retired military officer gave at another venue (while searching for the data, there were a few instances in which newspapers published remarks of officers made at events).

@@

Actually obtaining the data is a time consuming process. I use a series of key-word searches, such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "General," "Admiral," and so forth in several different databases, including Factiva, Lexus-Uni, and ProQuest Historical Newspapers. ^[For exact replication of the database search parameters, please see the Appendix.] Each of these databases, I learn, classifies historical newspaper articles differently: some databases include editorials as a separate source type to search for (making the acquisition of data easier), while others do not, requiring a search of all articles that contain the key words (which therefore results in a significantly larger pool of articles to search through and find the op-eds and letters I am looking for). I include in my search publications that were made available both in print and online format, though I did not double-count: if the same op-ed was published both in print and online and published by the same author, I count the article only one time. ^[To my knowledge, there is no data set that yet includes the data assembled for this research project. Therefore, I cannot conclusively say that I actually found all of the opinion pieces and letters authored by retired military officers and published in these sources within the time periods examined in this paper. It is possible, for instance, that I left out an important search parameter that did not trigger a certain number of op-eds that were actually authored by retired military officers. While it is difficult to put a precise degree of confidence in the method employed to obtain the data, my personal reaction is that the method employed caught greater than 95% of the actual observations. I offer this as a "ballpark" assessment based off several searches of individual newspapers in a given year in which I compared the results I obtained using the search parameters described against a method in which I literally looked through every op-ed published in a given calendar year.]

In some instances, especially where the same retired military officers authored multiple publications, the officer did not always refer to him or herself as a retired military officer. This occurred in several instances in which the officer was, ostensibly, writing from the perspective of someone other than a retired military officer, such as that of a political appointee to a high level governmental position (after retirement from the military, Army General Barry McCaffrey, for instance, served as President Clinton's Director of the Office of National Drug Control Policy). Nonethless, because these officers had previously identified themselves as retired military officers in earlier publications, and thus, because members of the public might reasonably recognize the author as a prominent retired military figure, I include the subsequent publications in the data, and simply control for whether the author identified him or herself as a retired military officer in a specific publication.

>>>>

@@

```{r, include = FALSE, echo = FALSE}
library("here")       # file paths
library("tidyverse")  # workhorse package
library("tidylog")
library("kableExtra")
library("knitr")
library("ggdag")
library("dagitty")
library("gridExtra")
library("broom")
library("dplyr")
library("reshape2")
library("stargazer")
library("rstatix")
library("lme4")
library("brglm")
library("bucky")
```

@@

This chapter examines variation in the publication of op-ed pieces by retired military officers since 1979. There are two main arguments of this chapter. The first is that the data clearly demonstrates a significant change in the number, topic, and tone of opinion-editorial publications published by retired military officers. The second argument holds that the primary driver of this change is the increase in political polarization that has occurred in America over the same time period. While other factors certainly play a role in influencing retired military officers to pick up the pen and write pieces that challenge various norms of officership, such as non-partisanship and civilian control of the military, the data clearly reveals that political polarization holds more explanatory power than other factors in explaining the variation of retired military officer speech in America over the past four decades. Together, this analysis highlights the extent to which retired military officer speech has strayed from professional normative standards.

This paper proceeds in three parts. First, the paper connects the behavior of retired military officer speech to the larger concept of military "role beliefs," a concept that is, I argue, worthy of reflection today, when both political polarization and the degree of societal prestige afforded to the military are at relatively high levels. Second, the paper explores and analyzes an original data set of nearly 390 opinion-editorial pieces published in major US newspapers by retired military officers since the late 1970s, and concludes that although the vast majority of op-eds published by retired military officers maintain professional standards, there is clear evidence that retired military officers are using the pen both to argue about topics that fall outside of the realm of traditional national security topics, and to violate norms of civilian control of the military and non-partisanship when doing so. Third, this paper discusses the implications of these findings on the challenges that face the military in what is perhaps the most difficult challenge it has faced in a long time: remaining professional in a highly polarized and divided society.

\singlespace
# How the Military Views Itself
\doublespace
## Military Role Beliefs 
The concept of military "role beliefs" was first introduced by J. Samuel Fitch in his book, _The Armed Forces and Democracy in Latin America._ In the book, Fitch describes military "role beliefs" as the perceptions held by military officers regarding their normative role in politics [@fitch_armed_1998, 61]. At the center of the concept of role beliefs are the ideas that military officers have about the extent to which they should participate in various aspects of the political process and engage in various forms of political behavior. While Fitch's focus is on militaries in Latin America, it is possible to use the concept of "role beliefs" to develop a normative starting point applicable to military officers in the United States.

Theoretically, there are at least two major determinants of military role beliefs, or the conceptions held by members of the military regarding their normative role in politics [@fitch_armed_1998]. In the most general terms, the military's involvement in politics requires both the _motive_ and the _opportunity_ to do so [@finer_man_1962]. I argue that in today's domestic political environment,  rising political polarization serves as the _motive_ for the military's intervention in politics, while high levels of prestige, or credibility of the military institution, provides the _opportunity_ for doing so. It is important to briefly elaborate how each of these offers the military in particular the ability to intervene in politics.

With respect to the first determinant, political polarization, the literature in American politics enables the development of two plausible relationships and corresponding mechanisms that connect polarization to  military role beliefs. These relationships and mechanisms further stem from a particular conception or definition of what polarization is and how it works. One possible conception of polarization is that it marks the ideological widening between political elites, to include members of Congress or the leaders of political parties [@hare_polarization_2014]. According to this view of polarization, one possible mechanism by which the military intervenes in politics is by adopting a belief that elites from both dominant political parties have failed to advance and implement an effective governing agenda, and thus, the military steps in to do what political elites are not doing. This particular dynamic has been noted, in fact, by scholars in the comparative politics sub-field describing the actions of militaries in other parts of the world [@fitch_armed_1998].

## Civilian Control, and Non-Partisanship

There are at two seminal principles that have helped guide and influence the military's "role beliefs" for most of US history: the tandem norms of civilian control of the military, and the non-partisanship of the military as an institution. Though these norms have been episodically challenged in a variety of specific instances and events, both norms have prominently featured within the military's organizational culture, and increasingly so as the military has modeled the behaviors of a profession, particularly in the post World War Two period [@weigley_american_2001]. The first norm of civilian control of the military stems directly from several provisions outlined in the US Constitution, which expressly grants control of the military (as well as a division of specific roles and responsibilities associated with maintaining control) to the President, Congress, and the individual states [@noauthor_us_1787-2]. The second norm of non-partisanship, while not expressly spelled out in the Constitution, has been cultivated throughout the military's collective experience of fighting and prevailing in the nation's wars and conflicts.

A number of scholars have likewise underscored the importance of and relationship between the norms of civilian control and non-partisanship of the military. Perhaps the most influential of these scholars has been Samuel Huntington, whose book, _The Soldier and the State: The Theory and Politics of Civil-Military Relations_, argued that maintaining the norms of civilian control of the military and non-partisanship would also ensure that the military prevailed against the dangerous threat environment of the Cold War. Huntington specifically argues in favor of a model of civilian control of the military known as "objective control" in which the military and civilian occupy largely separate spheres of operation such that civilian leaders do not interfere with the details of military operations, and likewise, the military does not intervene in politics [@huntington_soldier_1957]. Only by such an arrangement, Huntington argues, would the military exist as a "politically sterile and neutral," and thus, non-partisan, institution, while simultaneously ensuring civilian control of the military [@huntington_soldier_1957, 83-84].

## Opinion-Editorial Publication as a Political Activity?

Scholars have long commented on where the line should be drawn that separates appropriate vs. inappropriate political behavior by military officers, to include military retirees [@brooks_militaries_2009; @owens_military_2015; @golby_beyond_2015]. Most scholars note that this is in many ways a balancing act: while the norms of civilian control of the military and non-partisanship are as close to what can reasonably be considered "sacred" within and among the military, military officers are (especially in the case of those who achieve high rank) have accumulated unique knowledge and experiences over the course of lengthy careers (in the case of retired military officers, at least 20 and in some cases, upwards of 40 or even more years of service). In many cases, these military elites have interacted with Presidents, members of the cabinet, members of Congress, and foreign dignitaries. Though they possess an abundance of expertise in many tactical matters that pertain to war-fighting, they are also familiar with issues ranging from how wars should be fought, to future security threats looming on the horizon, to fixing the institutional culture of large organizations such as the military services, and much more. Moreover, especially in a democracy such as the United States, there is a reasonable expectation that these leaders use their military experience to contribute to the free flow of information that helps inform the citizens of a Republic and enables elected leaders to weigh the decisions before them. Nonetheless, military elites are not experts in everything, nor should they try to be. Even General of the Army Douglas MacArthur, one of several men who ever wore the five star rank, and a hero of World War Two who later clashed with President Truman regarding American strategic goals in Asia, recognized that the military should have focus areas. Addressing the Corps of Cadets at West Point in 1962, the then-82 year old retired General told the Corps:

@@

There have likewise been both formal and informal boundaries placed on various forms of political activities by military officers over the years. Formally, Department of Defense (DoD) Directive 1344.10 contains a "standing list of political behavior do's and don'ts for members of the active duty military" [@urben_wearing_2014, 1]. This list includes authorizing members of the military to engage in activities such as voting and joining a partisan or non-partisan club and to attend this club's meetings when not in uniform, and prohibits members of the military from activities such as participating in partisan fundraising, or advocating for or against a partisan cause or figure on radio or television [@department_of_defense_political_2008]. Informally, a number of recent Chairmen of the Joint Chiefs of Staff have likewise urged members of the military to refrain from engaging in overly political activities, particularly in election years, when such activity is more likely to negatively reflect on the military. For example, prior to the 2016 Presidential Election, then-Chairman of the Joint Chiefs of Staff General Joseph Dunford reminded the military that the basis upon which it is able to maintain the public's trust is it's ability to remain "apolitical and neutral" [@dunford_upholding_2016]. Clearly present in both the formal regulations and informal admonition regarding the political behavior of military officers is concern for the military's non-partisan ethos.

The specific act of authoring opinion commentary has occasionally garnered increased attention in recent years, perhaps starting in 1992 when then-Chairman of the Joint Chiefs of Staff General Colin Powell authored a piece that presented his views regarding the potential use of American military force in several developing "hot spots" across the world (including the Balkans, Somalia, and other places) entitled, "Why Generals Get Nervous." Published in the _New York Times_ less than a month away from the 1992 Presidential Election, Powell wrote in part, "decisive means and results are always to be preferred, even if they are not always possible. So you bet I get nervous when so-called experts suggest that all we need is a little surgical bombing or a limited attack. When the desired result isn't obtained, a new set of experts then comes forward with talk of a little escalation. History has not been kind to this approach" [@powell_opinion_1992]. More than a decade later, in 2006, the so-called "Revolt of the Generals" occurred in which several retired military officers publicly criticized and called for the removal of Defense Secretary Donald Rumsfeld as a result of his leadership and its impacts on the war in Iraq. Scholars noted that the event was unprecedented in the way that multiple officers "devoid of partisan political motive, felt the need or obligation to speak out publicly regarding policy and leadership during an ongoing conflict" [@cook_revolt_2008, 7]. While the "Revolt of the Generals" generated concern that the military's professional ethic had been strained, others applauded the generals' honesty and forthrightness and claimed that concerns about a dip in the professionalism among military officers was overblown. Frederick Kagan, for instance, claimed, "there is no danger to the republic in a handful of retired generals speaking their minds. There is great danger in making vital decisions about an on-going armed struggle without hearing the views of all available experts" [@kagan_let_2006].

Whether one agrees with Kagan's general assessment regarding the impact of opinion pieces authored by military officers, it is important to understand the ways in which opinion commentary is different from other types of speech made by military officers, such as that provided during formal testimony to Congress. Risa Brooks, for instance, argues that opinion or commentary is distinguished by both its "public nature" as well as the "scope and nature of the commentary" [@brooks_militaries_2009, 218]. Building off of Brooks, I would add that the action of authoring opinion commentary is completely voluntary, and thus, a deliberate action. Crafting a written op-ed is different from even giving interviews on television, because writing an op-ed requires the author to deliberately prepare and submit a viewpoint for publication. A deliberate and carefully-crafted message further underscores the degree to which the author believes the message he or she is sending. It is difficult to "misspeak" during or to be "misquoted" from an opinion-editorial.

In short, we should conclude that military officers who engage in authoring commentary or opinion pieces genuinely believes the message he or she is sending. As Brooks points out, however, the content of the specific message matters, and thus, the act of writing an op-ed in and of itself is not problematic. An op-ed that strictly points out the pitfalls of the nation's combat strategy while avoiding blatant criticism of a civilian official may not be inappropriate at all, whereas an op-ed written by a retired military officer that argues in favor of a specific Presidential candidate or policy is entirely inappropriate, at least by the normative standards of non-partisanship and civilian control of the military. Therefore, at least two questions should be asked when examining an opinion piece authored by a retired military officer: first, does the piece violate either the norm of civilian control of the military or the military as an apolitical institution, and second, does the op-ed address an appropriate topic that the military officer, on the basis of his or her professional expertise, is qualified to address?

## The Role of Retired Military Officers

Beyond these two questions that an individual officer (active or retired) should ask when thinking about writing an op-ed, there is, I argue, an even larger and broader question: what message or messages is the Republic receiving from the collective body of military leaders who provide opinion-laden commentary? Because military elites tend to provide this type of commentary more often than those on active duty, this question is particularly relevant for retired military elites. The collective tone, nature, and message sent by retired military elites serves as a barometer of sorts regarding the role that the retired officer corps sees for itself in interacting with the American public. Are retired military elites upholding norms of non-partisanship and civilian control of the military? When they are not, should we view these instances as a rare aberration, or are there clear trends that can be detected that should legitimately cause concern?

One reason to believe that a shift in the "role beliefs" among retired military officers may be occurring is the increasing political behavior of retired military elites during recent national political events. During the Presidential Campaign of 2016, for instance, Retired Army General Martin Dempsey, who previously served as Chairman of the Joint Chiefs of Staff, published an opinion-editorial piece in _The Washington Post_ which criticized retired Army Major General Mike Flynn and Marine Corps General John Allen for participating in the respective Republican and Democratic conventions. Dempsey wrote, "As generals, they have an obligation to uphold our apolitical traditions. They have just made the task of their successors – who continue to serve in uniform and are accountable for our security - more complicated. It was a mistake for them to participate as they did. It was a mistake for our presidential candidates to ask them to do so" [@dempsey_military_2016]. Thus, there was both a clear expectation and a concern among some prominent military elites that their fellow retired generals and admirals were not adhering to the norms of civilian control and non-partisanship of the military institution. Additionally, after four years of the Trump administration, during which several retired military officers served in prominent positions within the administration, and now that President-Elect Biden has, at least as of the time of writing this paper, nominated another retired general for the position of Secretary of Defense, there is a renewed interest in the normative roles that retired military elites should and actually do have in America.

By systematically examining the op-eds written by retired military officers in major US publications over time, I contend that it is possible to identify potential changes (or none) in the role beliefs held by retired US military officers. In doing so, the key area of focus must be the degree to which retired military officers have upheld vice strayed from the norms of civilian control and non-partisanship of the military, as well as the degree to which retired officers have addressed topics that are appropriately rooted in relevant military expertise. If there are patterns of violating these norms that can be detected, it may indicate that the role beliefs of certain members of the retired military community have likewise changed. On the other hand, if there has been little deviation over time from these foundational norms, it is possible to conclude that military retirees view their normative roles in politics (role beliefs) in similar ways as previous generations of retired military officers.

@@

Given the theoretical importance of the norms of civilian control and non-partisanship of the military institution, the rise of domestic political polarization and prestige of the US military over the past several decades, and the unique contributions that retired military elites make in a democratic society, four distinct hypotheses are derived, which will be examined and tested in the analysis to follow. These hypotheses include:

\singlespace
**H1: Increasing political polarization is associated with an increase in commentary authored by retired military officers that criticizes civilian leadership or makes clear partisan statements (potentially indicating a drop in adherence among retired officers to the principle of civilian control).**

@@

\singlespace
**H2: Increasing political polarization is associated with an increase in commentary authored by retired military officers that addresses topics that are outside of traditional national security debates.**

@@

\singlespace
**H3: During periods of high prestige and credibility, the aggregate level of commentary offered by retired military officers is higher relative to lower periods of prestige and credibility.**

@@

\singlespace
**H4: During periods of active ground combat, military elites will increasingly comment on topics related to the war, and comment less on other domestic issues.**

@@

\singlespace

@@

@@

The methodological approach used in this analysis consists of two broad steps. The first is to analyze the topics which retired military officers have been writing about, and the second is to determine, regardless of the topic addressed in a publication, whether an op-ed has violated the norms of non-partisanship or civilian control of the military. The data consists of 390 opinion-editorial publications and letters to the editor authored by retired military officers and published in major US newspapers from 1979-2020. I include both opinion-editorials and letters to the editor because both display a viewpoint that required the author to craft a deliberate message, and to submit the piece for consideration to be published by an editorial board at a particular newspaper. Because I am interested in messages that required the officer to craft a deliberate message, I do not include published interviews with retired military officers, nor any reprinted speeches or remarks that a retired military officer gave at another venue (while searching for the data, there were a few instances in which newspapers published remarks of officers made at events). I also do not include short forms of analysis, such as that provided by the "Monkey Cage" from _The Washington Post_.

@@

To obtain the data, I use a series of key-word searches, such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "General," "Admiral," and so forth in several different databases, including Factiva, Lexus-Uni, and ProQuest Historical Newspapers. ^[For exact replication of the database search parameters, please see the Appendix.] Each of these databases classifies historical newspaper articles differently: some databases include editorials as a separate source type to search for (making the acquisition of data easier), while others do not, requiring a search of all articles that contain the key words (which therefore results in a significantly larger pool of articles to search through and find the op-eds and letters I am looking for). I include in my search publications that were made available both in print and online format, though I did not double-count: if the same op-ed was published both in print and online and published by the same author, I count the article only one time. ^[To my knowledge, there is no data set that yet includes the data assembled for this research project. Therefore, I cannot conclusively say that I actually found all of the opinion pieces and letters authored by retired military officers and published in these sources within the time periods examined in this paper. It is possible, for instance, that I left out an important search parameter that did not trigger a certain number of op-eds that were actually authored by retired military officers. While it is difficult to put a precise degree of confidence in the method employed to obtain the data, my personal reaction is that the method employed caught greater than 95% of the actual observations. I offer this as a "ballpark" assessment based off several searches of individual newspapers in a given year in which I compared the results I obtained using the search parameters described against a method in which I literally looked through every op-ed published in a given calendar year.]

In some instances, especially where the same retired military officers authored multiple publications, the officer did not always refer to him or herself as a retired military officer. This occurred in several instances in which the officer was, ostensibly, writing from the perspective of someone other than a retired military officer, such as that of a political appointee to a high level governmental position (after retirement from the military, Army General Barry McCaffrey, for instance, served as President Clinton's Director of the Office of National Drug Control Policy). Nonetheless, because these officers had previously identified themselves as retired military officers in earlier publications, and thus, because members of the public might reasonably recognize the author as a prominent retired military figure, I include the subsequent publications in the data, and simply control for whether the author identified him or herself as a retired military officer in a specific publication.
